### 2020-06-08 Model Data Sheets and Code/Zaca MCF Mortality Model Results.Rmd
worked on rfm model and added ndvi to the data sheet

### Load data and packages
```{r}
library(randomForest)
library(pdp)
library(caret)
library(tidyverse)
library(ggpubr)
library(caret)
library(stringr)
library(here)
library(janitor)
library(ggfortify)
library(ALEPlot)
library(iml)

mcf <- read.csv('Zaca MCF variables for model.csv') %>% 
  select('drainage_association', 'ridge', 'mortality', 
         'prefire','mortality_qual', 'canopy_loss_success', 
         'canopy_loss_failure', 'postfire_canopy', 'postfire', 
         'Change_in_canopy_success', 'Change_in_canopy_failure', 
         'Change_Qualitative','high_mortality', 'area_scaled', 
         'dem_10m_zaca', 'flow_accumulation', 'sr', 'sr_new', 
         'rain', 'slope', 'tmax', 'tmean', 'tmin', 'TPI_five', 
         'TPI_queen', 'TWI', 'fri', 'dem_scaled', 'southness', "cwd", "vpd_jan", "vpd_aug", "tc_mortality" ) 
 #mutate(mortality_qual= fct_relevel(mortality_qual, levels=c("None", "Low", "Moderate", "High")))
 # keep getting error (Outer names are only allowed for unnamed scalar atomic inputs) when order the levels, but the reordering works anyway (although it doesnt work with the ssame error in the binomial code. The code directly below works to reorder the levels without error)
mcf$mortality_qual  <- factor(mcf$mortality_qual, levels = c("None", "Low", "Moderate", "High"))

@@

@@

```{r}
solar_rad_mcf <- ggplot(mcf, aes(x=mortality_qual, y=sr, fill=mortality_qual))+
  geom_violin(alpha=0.5, show.legend=F) +
  theme_classic() +
  labs(x="Mortality", y="Solar Radiation") +
  scale_fill_manual(values=c("red", "orange", "yellow", "green")) +
  coord_flip()

tmax_mcf <- ggplot(mcf, aes(x=mortality_qual, y=tmax, fill=mortality_qual))+
  geom_violin(alpha=0.5, show.legend=F) +
  theme_classic() +
  labs(x="Mortality", y="Max Aug Temp (celcius)") +
  scale_fill_manual(values=c("red", "orange", "yellow", "green")) +
  coord_flip()

rain_mcf <- ggplot(mcf, aes(x=mortality_qual, y=rain, fill=mortality_qual))+
  geom_violin(alpha=0.5, show.legend=F) +
  theme_classic() +
  labs(x="Mortality", y="Precipitation") +
  scale_fill_manual(values=c("red", "orange", "yellow", "green")) +
  coord_flip()

tpi_mcf <- ggplot(mcf, aes(x=mortality_qual, y=TPI_five, fill=mortality_qual))+
  geom_violin(alpha=0.5, show.legend=F) +
  theme_classic() +
  labs(x="Mortality", y="TPI") +
  scale_fill_manual(values=c("red", "orange", "yellow", "green")) +
  coord_flip()

twi_mcf <- ggplot(mcf, aes(x=mortality_qual, y=TWI, fill=mortality_qual))+
  geom_violin(alpha=0.5, show.legend=F) +
  theme_classic() +
  labs(x="Mortality", y="TWI") +
  scale_fill_manual(values=c("red", "orange", "yellow", "green")) +
  coord_flip()

slope_mcf <- ggplot(mcf, aes(x=mortality_qual, y=slope, fill=mortality_qual))+
  geom_violin(alpha=0.5, show.legend=F) +
  theme_classic() +
  labs(x="Mortality", y="Slope") +
  scale_fill_manual(values=c("red", "orange", "yellow", "green")) +
  coord_flip()

elev_mcf <- ggplot(mcf, aes(x=mortality_qual, y=dem_10m_zaca, fill=mortality_qual))+
  geom_violin(alpha=0.5, show.legend=F) +
  theme_classic() +
  labs(x="Mortality", y="Elevation") +
  scale_fill_manual(values=c("red", "orange", "yellow", "green")) +
  coord_flip()

cwd_mcf <- ggplot(mcf, aes(x=mortality_qual, y=cwd, fill=mortality_qual))+
  geom_violin(alpha=0.5, show.legend=F) +
  theme_classic() +
  labs(x="Mortality", y="CWD") +
  scale_fill_manual(values=c("red", "orange", "yellow", "green")) +
  coord_flip()

aug_vpd_mcf <- ggplot(mcf, aes(x=mortality_qual, y=vpd_aug, fill=mortality_qual))+
  geom_violin(alpha=0.5, show.legend=F) +
  theme_classic() +
  labs(x="Mortality", y="Aug VPD") +
  scale_fill_manual(values=c("red", "orange", "yellow", "green")) +
  coord_flip()

jan_vpd_mcf <- ggplot(mcf, aes(x=mortality_qual, y=vpd_jan, fill=mortality_qual))+
  geom_violin(alpha=0.5, show.legend=F) +
  theme_classic() +
  labs(x="Mortality", y="Jan VPD") +
  scale_fill_manual(values=c("red", "orange", "yellow", "green")) +
  coord_flip()

@@

```{r}
mcf_sub <- mcf %>% 
  select (dem_10m_zaca:sr, rain:tmax, TPI_five, TWI, cwd:vpd_aug) %>% 
  rename ('Elevation' = 'dem_10m_zaca',
          'Slope' = 'slope',
          'TPI' = 'TPI_five',
          'Precip' = 'rain',
          'Solar Radiation' = 'sr',
          'Flow Accumulation' = 'flow_accumulation',
          'CWD' = 'cwd',
          'Temperature' = 'tmax',
          'VPD (Jan)' = 'vpd_jan',
          'VPD (Aug)' = 'vpd_aug')

@@

@@

  #random forest model
  rfm<-randomForest(mortality_qual~ rain + tmax + TPI_five + dem_10m_zaca + flow_accumulation + sr + slope + TWI + cwd + vpd_aug, data = train, ntree=500, importance=T, nscale=0)

@@

### Random Forest Model with 3 mortality classes

@@

```{r}
# model parameters and sub data -----------------------------------------------------------

# sub data 
mcf_3 <- mcf %>%
  mutate(three_class_mortality = case_when(
    mortality_qual %in% "High" ~ "High",
    mortality_qual %in% "Moderate" ~ "Moderate",
    mortality_qual %in% c("Low", "None") ~ "Low"
  )) %>% 
  mutate(three_class_mortality = as.factor(three_class_mortality)) %>% 
  mutate(three_class_mortality= fct_relevel(three_class_mortality, 
                                     levels=c("Low", "Moderate", "High")))

@@

@@

  mcf_high <- mcf_3[which(mcf_3$three_class_mortality == "High"),]
  mcf_low <- mcf_3[which(mcf_3$three_class_mortality == "Low"),]
  mcf_moderate <- mcf_3[which(mcf_3$three_class_mortality == "Moderate"),]

@@

  #random forest model
  rfm<-randomForest(three_class_mortality~ rain + tmax + TPI_five + dem_10m_zaca + flow_accumulation + sr + slope + TWI + cwd + vpd_aug, data = train, ntree=500, importance=T, nscale=0)

  #plot to visualize variable importance
  model_importance <- importance(rfm)

  #varImpPlot(rfm)

@@

  #random forest model
  rfm<-randomForest(tc_mortality~ rain + tmax + TPI_five + dem_10m_zaca + flow_accumulation + sr + slope + TWI + cwd + vpd_aug + ridge + drainage_association, data = train, ntree=500, importance=T, nscale=0)

@@

@@

@@

# sub-data: combine low and no mortality classes
mcf_3 <- mcf %>%
  mutate(mortality_qual_3 = case_when(
    mortality_qual %in% c("None", "Low") ~ "Low",
    mortality_qual == "Moderate" ~ "Moderate",
    mortality_qual == "High" ~ "High"
  )) %>% 
  mutate(canopy_loss_success_3 = case_when(
    canopy_loss_success %in% c("0", "250") ~ "250",
    canopy_loss_success == "500" ~"500",
    canopy_loss_success == "750" ~ "750"
  )) %>% 
  mutate(canopy_loss_failure_3 = case_when(
    canopy_loss_failure %in% c("1000", "750") ~ "750",
    canopy_loss_failure == "500" ~"500",
    canopy_loss_failure == "250" ~ "250"
  ))

mcf_3$canopy_loss_success_3 <- factor(mcf_3$canopy_loss_success_3)
mcf_3$canopy_loss_failure_3 <- factor(mcf_3$canopy_loss_failure_3)

@@

for (i in 1:iteration1) {
# train and test data 
  mcf_high <- mcf_3[which(mcf_3 $ mortality_qual_3 == "High"),]
  mcf_low <- mcf_3[which(mcf_3 $ mortality_qual_3 == "Low"),]
  mcf_moderate <- mcf_3[which(mcf_3 $ mortality_qual_3 == "Moderate"),]

>>>>

### Load data and packages
```{r}
library(randomForest)
library(pdp)
library(caret)
library(tidyverse)
library(ggpubr)
library(caret)
library(stringr)
library(here)
library(janitor)
library(ggfortify)
library(ALEPlot)
library(iml)
library(ROCR)
library(pROC)

mcf <- read.csv('Zaca MCF variables for model.csv') %>% 
  mutate(mortality_qual= fct_relevel(mortality_qual, levels=c("None", "Low", "Moderate", "High"))) %>% 
  # mutate(dnbr_qualitative = case_when(
  #   dnbr < 0.1 ~ "unburned",
  #   dnbr < 0.27 ~ "low",
  #   dnbr < 0.66 ~ "moderate",
  #   dnbr > 0.66 ~ "high")) %>% 
  # mutate(dnbr_qualitative = as.factor(dnbr_qualitative),
  #        dnbr_qualitative = fct_relevel(dnbr_qualitative, levels=c("high", "moderate", "low", "unburned"))) %>% 
  mutate(aspect_transformed = (1-cos((2*pi*aspect)/360))/2) %>% 
  mutate(three_class_mortality = case_when(
    mortality_qual %in% "High" ~ "High",
    mortality_qual %in% "Moderate" ~ "Moderate",
    mortality_qual %in% c("Low", "None") ~ "Low")) %>% 
  mutate(three_class_mortality = as.factor(three_class_mortality)) %>% 
  mutate(three_class_mortality= fct_relevel(three_class_mortality, 
                                     levels=c("Low", "Moderate", "High")))

 #mutate(mortality_qual= fct_relevel(mortality_qual, levels=c("None", "Low", "Moderate", "High")))
 # keep getting error (Outer names are only allowed for unnamed scalar atomic inputs) when order the levels, but the reordering works anyway (although it doesnt work with the ssame error in the binomial code. The code directly below works to reorder the levels without error)

```
for dnbr classes see: https://www.int-arch-photogramm-remote-sens-spatial-inf-sci.net/XLI-B8/763/2016/isprs-archives-XLI-B8-763-2016.pdf

for aspect transformation: https://esajournals.onlinelibrary.wiley.com/doi/full/10.1890/07-0539.1 appendix B; transformed Aspect = [1 - cos(2piAspect/360)]/2. This formula is similar (but not identical) to that used by Roberts and Cooper (1989). The transformed aspect values lie in the interval from 0 to1. Values near 0 represent aspects close to due north, while values near 1 represent aspects close to due south. East and west are treated identically with transformed aspect values of 0.5.

@@

tc_count <- mcf %>% 
  group_by(tc_mortality) %>% 
  count() # 99 stands with type conversion, 636 with no type conversion

```
None: 251
Low: 166
mod: 160
high: 158

@@

```{r}
solar_rad_mcf <- ggplot(mcf, aes(x=mortality_qual, y=sr, fill=mortality_qual))+
  geom_violin(alpha=0.5, show.legend=F) +
  theme_classic() +
  labs(x="Mortality", y="Solar Radiation") +
  scale_fill_manual(values=c("green", "yellow", "orange", "red")) +
  coord_flip()

tmax_mcf <- ggplot(mcf, aes(x=mortality_qual, y=tmax, fill=mortality_qual))+
  geom_violin(alpha=0.5, show.legend=F) +
  theme_classic() +
  labs(x="Mortality", y="Max Aug Temp (celcius)") +
  scale_fill_manual(values=c("green", "yellow", "orange", "red")) +
  coord_flip()

rain_mcf <- ggplot(mcf, aes(x=mortality_qual, y=rain, fill=mortality_qual))+
  geom_violin(alpha=0.5, show.legend=F) +
  theme_classic() +
  labs(x="Mortality", y="Precipitation") +
  scale_fill_manual(values=c("green", "yellow", "orange", "red")) +
  coord_flip()

tpi_mcf <- ggplot(mcf, aes(x=mortality_qual, y=TPI_five, fill=mortality_qual))+
  geom_violin(alpha=0.5, show.legend=F) +
  theme_classic() +
  labs(x="Mortality", y="TPI") +
  scale_fill_manual(values=c("green", "yellow", "orange", "red")) +
  coord_flip()

twi_mcf <- ggplot(mcf, aes(x=mortality_qual, y=TWI, fill=mortality_qual))+
  geom_violin(alpha=0.5, show.legend=F) +
  theme_classic() +
  labs(x="Mortality", y="TWI") +
  scale_fill_manual(values=c("green", "yellow", "orange", "red")) +
  coord_flip()

slope_mcf <- ggplot(mcf, aes(x=mortality_qual, y=slope, fill=mortality_qual))+
  geom_violin(alpha=0.5, show.legend=F) +
  theme_classic() +
  labs(x="Mortality", y="Slope") +
  scale_fill_manual(values=c("green", "yellow", "orange", "red")) +
  coord_flip()

elev_mcf <- ggplot(mcf, aes(x=mortality_qual, y=dem_10m_zaca, fill=mortality_qual))+
  geom_violin(alpha=0.5, show.legend=F) +
  theme_classic() +
  labs(x="Mortality", y="Elevation") +
  scale_fill_manual(values=c("green", "yellow", "orange", "red")) +
  coord_flip()

cwd_mcf <- ggplot(mcf, aes(x=mortality_qual, y=cwd, fill=mortality_qual))+
  geom_violin(alpha=0.5, show.legend=F) +
  theme_classic() +
  labs(x="Mortality", y="CWD") +
  scale_fill_manual(values=c("green", "yellow", "orange", "red")) +
  coord_flip()

aug_vpd_mcf <- ggplot(mcf, aes(x=mortality_qual, y=vpd_aug, fill=mortality_qual))+
  geom_violin(alpha=0.5, show.legend=F) +
  theme_classic() +
  labs(x="Mortality", y="Aug VPD") +
  scale_fill_manual(values=c("green", "yellow", "orange", "red")) +
  coord_flip()

jan_vpd_mcf <- ggplot(mcf, aes(x=mortality_qual, y=vpd_jan, fill=mortality_qual))+
  geom_violin(alpha=0.5, show.legend=F) +
  theme_classic() +
  labs(x="Mortality", y="Jan VPD") +
  scale_fill_manual(values=c("green", "yellow", "orange", "red")) +
  coord_flip()

ggplot(mcf, aes(x=mortality_qual, y=ndvi, fill=mortality_qual))+
  geom_violin(alpha=0.5, show.legend=F) +
  theme_classic() +
  labs(x="Mortality", y="NDVI") +
  scale_fill_manual(values=c("green", "yellow", "orange", "red")) +
  coord_flip()

@@

```{r}
mcf_sub <- mcf %>% 
  select (dem_10m_zaca:sr, rain:tmax, TPI_five, TWI, cwd:aspect_transformed) %>% 
  rename ('Elevation' = 'dem_10m_zaca',
          'Slope' = 'slope',
          'TPI' = 'TPI_five',
          'Precip' = 'rain',
          'Solar Radiation' = 'sr',
          'Flow Accumulation' = 'flow_accumulation',
          'CWD' = 'cwd',
          'Temperature' = 'tmax',
          'VPD (Jan)' = 'vpd_jan',
          'VPD (Aug)' = 'vpd_aug')

@@

### Random Forest: ALL observations

to get a stable variable importance. The other models result in very unstable variable importance cmeasures which shuldnt happen with RF but is occuring bc the training data used for each run changes.

```{r}
#four class mortality
for (i in 1:10) {

rfm <- randomForest(mortality_qual ~ rain + tmax + TPI_five + dem_10m_zaca + flow_accumulation + sr + slope + TWI + cwd + vpd_aug + ndvi, data = mcf, ntree=1500, importance=T, nscale=0)

rfm

#plot to visualize variable importance
varImpPlot(rfm)
model_importance <- importance(rfm)

#3 class mortality
for (i in 1:10) {

rfm <- randomForest(three_class_mortality ~ rain + tmax + TPI_five + dem_10m_zaca + flow_accumulation + sr + slope + TWI + cwd + vpd_aug + ndvi, data = mcf, ntree=1500, importance=T, nscale=0)

rfm

#plot to visualize variable importance
varImpPlot(rfm)
model_importance <- importance(rfm)

```
So 1500 trees reuslts in more stable importance values for each model. The model with 3 mortality classes has a better OOB.

# Random Forest: 4 mortality classes
```{r}
# model parameters---------------------------------------------

@@

  #random forest model
  rfm<-randomForest(mortality_qual~ rain + tmax + TPI_five + dem_10m_zaca + flow_accumulation + sr + slope + TWI + cwd + vpd_aug + ndvi, data = train, ntree=500, importance=T, nscale=0)

@@

### Random Forest Model: 3 mortality classes

@@

```{r}
# model parameters ------------------------------

@@

#random forest code---------------------------------------------------
for (i in 1:5)
{

@@

  mcf_high <- mcf[which(mcf$three_class_mortality == "High"),]
  mcf_low <- mcf[which(mcf$three_class_mortality == "Low"),]
  mcf_moderate <- mcf[which(mcf$three_class_mortality == "Moderate"),]

@@

  #random forest model
  rfm<-randomForest(three_class_mortality~ rain + tmax + TPI_five + dem_10m_zaca + ndvi + sr + slope + TWI + cwd + vpd_aug, data = train, ntree=500, importance=T, nscale=0)

  #plot to visualize variable importance
  model_importance <- importance(rfm)
  varImpPlot(rfm)

@@

  #random forest model
  rfm <- randomForest(tc_mortality ~ rain + tmax + TPI_five + dem_10m_zaca + flow_accumulation + sr + slope + TWI + cwd + vpd_aug + ridge + drainage_association + ndvi, data = train, ntree=1500, importance=T, nscale=0)

  #plot to visualize variable importance
  varImpPlot(rfm)
  model_importance <- importance(rfm)

@@

accuracy_df <- as.data.frame(unlist(test_accuracy_list))
mean_model_accuracy_1 <- accuracy_df %>% 
  summarize(mean_accuracy = mean(unlist(test_accuracy_list)))

@@

#### ROC and AUC
```{r}
# code: https://stats.stackexchange.com/questions/188616/how-can-we-calculate-roc-auc-for-classification-algorithm-such-as-random-forest
rfm_roc <-  roc(train$tc_mortality, rfm$votes[,1])
plot(rfm_roc) # plot comes out a little weird
auc(rfm_roc)

@@

# sub-data: combine low and no mortality classes
mcf <- mcf %>%
  mutate(mortality_qual_3 = case_when(
    mortality_qual %in% c("None", "Low") ~ "Low",
    mortality_qual == "Moderate" ~ "Moderate",
    mortality_qual == "High" ~ "High"
  )) %>% 
  mutate(canopy_loss_success_3 = case_when(
    canopy_loss_success %in% c("0", "250") ~ "250",
    canopy_loss_success == "500" ~"500",
    canopy_loss_success == "750" ~ "750"
  )) %>% 
  mutate(canopy_loss_failure_3 = case_when(
    canopy_loss_failure %in% c("1000", "750") ~ "750",
    canopy_loss_failure == "500" ~"500",
    canopy_loss_failure == "250" ~ "250"
  ))

mcf$canopy_loss_success_3 <- factor(mcf$canopy_loss_success_3)
mcf$canopy_loss_failure_3 <- factor(mcf$canopy_loss_failure_3)

@@

for (i in 1:iteration1) {
# train and test data 
  mcf_high <- mcf[which(mcf $ mortality_qual_3 == "High"),]
  mcf_low <- mcf[which(mcf $ mortality_qual_3 == "Low"),]
  mcf_moderate <- mcf[which(mcf $ mortality_qual_3 == "Moderate"),]
